An Evolutionary Algorithm for the Multiobjective Risk-Equity Constrained Routing Problem

### 2011-11-29 trunk/DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
a bit more text (incrementally)

\subsection{Implementation in PISA}
The proposed algorithm has been implemented within the multiobjective optimization framework PISA \citep{bltz2003a} and is going to be made available via the official PISA web page\footnote{Until the source code is available via the PISA web page \url{http://www.tik.ee.ethz.ch/pisa/}, we made it temporarily available at \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip} as supplementary material for the review process.}.

@@

As a test case for all the experiments, we chose to use a real-world case study based on the work of \citet{bianco09}. The network used here consists of 311 nodes and 441 arcs and represents the main transport roads in Italy's Lazio region which, amongst other cities, includes the capital Rome. We distinguish between instances with a different number of commodities (between 2 and 4) and with different numbers of trucks, see Table~\ref{tab:instances}. While the costs $c_{ij}$ are proportional to the real distance between the two locations $i$ and $j$, the risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ have been computed as a product of the arc's length, the actual population density in the area of the arc and a constant which depends on the commodity, for details see \citep{bianco09} and Table~\ref{tab:instances}. Solely one instance is created differently: to show the impact of correlation between the costs and the risks, we created the instance lazio$_{\text ncorr}$\_4\_1 with the same graph and risks as for the instance lazio\_4\_1 but with randomly permutated costs where each occuring cost value has been divided by a random value between 0 and 5000.

\begin{table}
\caption{\label{tab:instances} Parameters of the instances used in the experimental analyses.}
\begin{tabular}{lllll}
	\toprule
	name & \begin{minipage}{1cm}\#com\-mod\-i\-ties\end{minipage} & \#trucks & cost range & risk range \\ \midrule
	lazio\_2\_1 & 2 & \small 20+19         & & \begin{minipage}[t]{3.7cm} \small [762, 274628]\\ x [610.18, 219702.6]\end{minipage}\\
	lazio\_2\_2 & 2 & \small 23+29         & & \\
	lazio\_3\_1 & 3 & \small 25+16+28      & & \begin{minipage}[t]{3.7cm} \small [762, 274628]\\ x [610.18, 219702.6]\\ x [1120.352, 439405.2]\end{minipage}\\
	lazio\_4\_1 & 4 & \small 21+3+30+5     & \small [0.25, 41] & \begin{minipage}[t]{3.7cm} \small [762, 274628]\\ x [610.18, 219702.6]\\ x [1120.352, 439405.2]\\ x [762, 274628]\end{minipage}\\
	lazio$_{\text uncorr}$\_4\_1 & 4 & \small 21+3+30+5     & \small [0.0109239, 3.93701] & \begin{minipage}[t]{3.7cm} \small [762, 274628]\\ x [610.18, 219702.6]\\ x [1120.352, 439405.2]\\ x [762, 274628]\end{minipage}\\
	\bottomrule
\end{tabular}
\end{table}

%We concentrated our analysis on a real-world case study \citep{bianco09}. We considered the road network of the Lazio region (located in the middle of Italy), and, in particular, its main transport roads. The network is composed of 311 nodes and 441 links. The risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ are evaluated as the societal risk computed as the number of people living inside the exposure zone around link $(i,j)$ (whose size depends on the hazmat type $c$ times the accident probability involving the vehicle), for details see  \citep{bianco09}. We considered from 2 to 10 origin-destination pairs on the network, each one associated with a number of commodities ranging from 1 to 3, for an overall number of shipments (commodities) between 2 and 30 (we recall that in our model each commodity is associated with one origin-destination pair). With each one of the latter scenarios we associated 10 instances; each instance has been generated by assigning uniformly at random a demand from 100 to 1000 tons to each shipment.

\subsection{Experimental Settings}
Except otherwise stated, we used the following parameter setting. We perform 10 independent runs of HypE for 2000 generations with a population size of 100. The reference point for the hypervolume indicator was chosen as in Table~\ref{tab:referencepoints} according to the initial objective function values, such that all initial solutions have a positive contribution to the hypervolume indicator\footnote{Note that the reason for using the same values in all objectives is the used PISA implementation of HypE.}. In all cases, the hypervolume computation is exact.

\begin{table}
\centering
\caption{\label{tab:referencepoints} Used reference points within the environmental selection of HypE}
\begin{tabular}{lr}
	\toprule
	instance & reference point\\ \midrule
	\# commodities: 2 & $(10^8, 10^8, 10^8)$\\
	\# commodities: 3 & $(10^8, 10^8, 10^8)$\\
	\# commodities: 4 & $(10^9, 10^9, 10^9)$\\
	\bottomrule
\end{tabular}
\end{table}

\subsection{Differences Between Instances}

@@

\subsection{Changing the Environmental Selection}
With a last experiment, we show the influence of the algorithm's environmental selection scheme on the performance and in particular how this can be used to steer the search towards specific regions of the Pareto front.
\TODO{write}

\subsection{The Feasibility of the EMO Approach}
\begin{itemize}
	\item of course no guarantee that the solutions are Pareto-optimal
	\item however, in reasonable time (e.g. ??hrs for 10 runs with 1000 generations each) results with several non-dominated solutions the decision maker can interpret and with which she can get an idea about the trade-offs between the objectives
	\item if DM wants good results fast, cost-efficient initialization is favourable, otherwise, random initialization gives better spread of the solutions
	\item if needed, approach can be combined easily with preference articulation techniques such as the weighted hypervolume indicator which allows to steer the search towards user-defined regions of the objective space (show 1 plot with W-HypE).
	\item approach is very general and can easily deal with slightly modified problems: e.g. different truck loads (only the objective function code needs to be changed)
	\item it needs to be mentioned that for the tested instances with 311 nodes and 441 arcs, the variation operator of the evolutionary algorithm was able to construct the shortest paths from source to destination in some of the runs for some of the truck paths within the first 100,000 function evaluations. This corresponds to performing random walks of lengths around 20 nodes.
\end{itemize}

>>>>

\subsection{Implementation in PISA}
The proposed algorithm has been implemented within the multiobjective optimization framework PISA \citep{bltz2003a} and is going to be made available via the official PISA web page\footnote{Until the source code is available via the PISA web page \url{http://www.tik.ee.ethz.ch/pisa/}, we made it temporarily available at \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip} as supplementary material for the review process.}. More precisely, the implementation of an algorithm/problem combination in PISA is divided into the problem-specific part (called \emph{variator module}) consisting of representation, initialization, and variation operators and the problem-independent part (called the \emph{selector module}). As selection module, any of the state-of-the-art selectors in PISA can be used off-the-shelf. In this study, we choose to use HypE \citep{bz2011a}, NSGA-II \citep{dapm2002a}, and W-HypE \citep{} with a focus on the first one.

After the initialization of the population with $\alpha$ individuals, those solutions are evaluated and their objective function values transferred to the selector module which then \TODO{write}

$\lambda$ individuals are chosen for variation and undergo the above described mutation operator creating the $\lambda$ so-called offspring.

 The objective vectors of the current population and these new solutions are then handed over to the selector module which decides which solutions to

\TODO{describe how PISA works}

@@

As a test case for all the experiments, we chose to use a real-world case study based on the work of \citet{bcg2009a}. The network used here consists of 311 nodes and 441 arcs and represents the main transport roads in Italy's Lazio region which, amongst other cities, includes the capital Rome. We distinguish between instances with a different number of commodities (between 2 and 4) and with different numbers of trucks, see Table~\ref{tab:instances}. While the costs $c_{ij}$ are proportional to the real distance between the two locations $i$ and $j$, the risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ have been computed as a product of the arc's length, the actual population density in the area of the arc and a constant which depends on the commodity, for details see \citep{bcg2009a} and Table~\ref{tab:instances}. Solely one instance is created differently: to show the impact of correlation between the costs and the risks, we created the instance uncorrelated\_4 with the same graph and risks as for the instance lazio\_4\_1 but with randomly permutated costs where each occuring cost value has been divided by a random value between 0 and 5000.

\begin{table}
\caption{\label{tab:instances} Parameters of the instances used in the experimental analyses with $|C|$ being the number of different commodities.}\vspace{1em}
\begin{tabular}{lllll}
	\toprule
	name & $|C|$ & \#trucks & cost range & risk range \\ \midrule
	lazio\_2 & 2 & \small 25+16         & \small [0.25, 41] & \begin{minipage}[c]{3.8cm} \small [762, 274628]\\ $\times$ [610.18, 219702.6]\end{minipage}\\
	\midrule
	lazio\_3 & 3 & \small 25+16+28      & \small [0.25, 41] & \begin{minipage}[c]{3.8cm} \small [762, 274628]\\ $\times$ [610.18, 219702.6]\\ $\times$ [1120.352, 439405.2]\end{minipage}\\
	\midrule
	lazio\_4\_1 & 4 & \small 21+3+30+5     &  & \multirow{4}{*}{\begin{minipage}[c]{3.8cm} \hspace{1pt}\\ \\ \small [762, 274628]\\ $\times$ [610.18, 219702.6]\\ $\times$ [1120.352, 439405.2]\\ $\times$ [762, 274628]\end{minipage}}\\
	\cmidrule{1 - 3}
	lazio\_4\_2 & 4 & \small 21+3+30+6     & \small [0.25, 41] & \\
	\cmidrule{1 - 3}
	lazio\_4\_3 & 4 & \small 7+1+10+2     &  & \\
	\cmidrule{1 - 4}
	uncorrelated\_4 & 4 & \small 21+3+30+5     & \begin{minipage}[t]{2cm} \small [0.0109239, 3.93701]  \end{minipage} & \\
	\bottomrule
\end{tabular}
\end{table}

%We concentrated our analysis on a real-world case study \citep{bcg2009a}. We considered the road network of the Lazio region (located in the middle of Italy), and, in particular, its main transport roads. The network is composed of 311 nodes and 441 links. The risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ are evaluated as the societal risk computed as the number of people living inside the exposure zone around link $(i,j)$ (whose size depends on the hazmat type $c$ times the accident probability involving the vehicle), for details see  \citep{bcg2009a}. We considered from 2 to 10 origin-destination pairs on the network, each one associated with a number of commodities ranging from 1 to 3, for an overall number of shipments (commodities) between 2 and 30 (we recall that in our model each commodity is associated with one origin-destination pair). With each one of the latter scenarios we associated 10 instances; each instance has been generated by assigning uniformly at random a demand from 100 to 1000 tons to each shipment.

\subsection{Experimental Settings}
Except otherwise stated, we used the following parameter setting. We perform 10 independent runs of HypE for 2000 generations with a population size of 50, i.e., each algorithm run has a budget of 100,000 function evaluations. The reference point for the hypervolume indicator was chosen as $(10^8, 10^8, 10^8)$ for the instances with 2 and 3 commodities and as $(10^9, 10^9, 10^9)$ for all instances with 4 commodities. The reference point has been chosen according to the objective function values of the random initial populations, such that all initial solutions have a positive contribution to the hypervolume indicator. Note that the reason for using the same values in all objectives is the used PISA implementation of HypE. In all cases, the hypervolume computation is exact.

\begin{table}
\centering
\caption{\label{tab:referencepoints} Used reference points within the environmental selection of HypE}
\vspace{1em}
\begin{tabular}{lr}
	\toprule
	instance & reference point\\ \midrule
	\# commodities: 2 & $(10^8, 10^8, 10^8)$\\
	\# commodities: 3 & $(10^8, 10^8, 10^8)$\\
	\# commodities: 4 & $(10^9, 10^9, 10^9)$\\
	\bottomrule
\end{tabular}
\end{table}

\subsection{Differences Between Instances}
The basis for our experiments are three instances of the lazio network with 2 (lazio\_2), 3 (lazio\_3), and 4 commodities (lazio\_4\_1) which differ also differ in the number of trucks for the commodities. Whereas the number of trucks for lazio\_3 and lazio\_4\_1 have been chosen randomly, lazio\_2 has the same number of trucks than the lazio\_3 instance for the first two commodities.

\TODO{show results and talk about them}

\begin{figure}%
	\centering
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_50.eps}%
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_100.eps}%
	\caption{\label{fig:hypervolumeOverTime} Boxplots of hypervolume values over time for numbers of generations in $[250, \ldots, 2000]$ for HypE with population size 50 (left) and population size 100 (right) on the lazio\_4 instance.}
\end{figure}

@@

\subsection{Changing the Environmental Selection}
With a last experiment, we show the influence of the algorithm's environmental selection scheme on the performance and in particular how this can be used to steer the search towards specific regions of the Pareto front.

Instead of the environmental selection of HypE \citep{bz2011a}, other selection schemes can be easily used with the PISA framework and we show the resutls of two other algorithms exemplary. On the one hand, we employ the famous NSGA-II selection scheme \citep{dapm2002a} and, on the other hand, we use a generalized version of HypE towards a generalized weighted hypervolume indicator \citep{abbz2009b,bbtz2011a}. The boxplots of the resulting hypervolume indicator values after 100,000 function evaluations (population of size 100 after 1000 generations) are shown in Fig.~\ref{fig:algoComparison}. \TODO{What can be seen is that both? algorithms show a reduced performance regarding the standard hypervolume value. However, W-HypE was run in order to show how the search can be steered towards regions of the objective space that have been specified by the user. In our case, we defined the weight of the underlying weighted hypervolume indicator \citep{zbt2007a} as a box with corners XXX such that the weighted hypervolume indicator only influences BLABLABLA}

\TODO{write}

\subsection{The Feasibility of the EMO Approach}
Due to their heuristic nature, evolutionary multiobjective optimization algorithms like the one presented in this study cannot guarantee that the solutions found are actually Pareto-optimal. However, as we have seen above, the proposed algorithm is able to produce a set of non-dominated solutions in reasonable time which are close to Pareto-optimal solutions found by exact algorithms. Such sets of non-dominated solutions can be a starting point for a decision maker to investigate the trade-offs between the objectives and to decide in which objective space region interesting Pareto-optimal solutions are which can be explored further with more involved algorithms---be it exact algorithms or specialized evolutionary algorithms such as the mentioned W-HypE variant. In the above example instances, it turned out that total risk and cost are indeed more or less fully correlated in the original problem formulation or only loosely (with a larger trade-off between cost and total risk) when the individual arcs' risks and costs are.

Regarding the two initialization schemes, it turned out that in combination with the (local and simple) variation operator, the cost-efficient initialization does not allow the algorithm to find additional solutions besides the initial one and that, in particular if the cost and total risk objectives are correlated, the random initialization is beneficial. Using other initializations as well as additional variation operators from the literature is easily possible with the proposed algorithm but their investigation is left for future research\footnote{For example, the

On the other hand, the proposed variation operator can be changed towards a larger neighborhood by allowing two or more nodes to be added or deleted from each truck path at a time---resulting in more exploration than exploitation when compared to the current variation operator. Also well-designed crossover operators are expected to be beneficial to the problem when, for example, good sub paths can be exchanged between solution paris.
}.

\begin{itemize}

	\item approach is very general and can easily deal with slightly modified problems: e.g. different truck loads (only the objective function code needs to be changed)
	\item it needs to be mentioned that for the tested instances with 311 nodes and 441 arcs, the variation operator of the evolutionary algorithm was able to construct the shortest paths from source to destination in some of the runs for some of the truck paths within the first 100,000 function evaluations. This corresponds to performing random walks of lengths around 20 nodes.
\end{itemize}
